---
title: "Convert flowSet to SingleCellExperiment with sample annotations"
subtitle: "High-dimensional cytometry workflow step 2"
date-modified: last-modified
format:
  html:
    embed-resources: true
    code-tools: true
    toc: true
    df-print: paged
execute:
  cache: false
params:
  # project label for the input/output directory
  project_name: "project"
  # cofactor for arcsinh transformation
  # generally 5 for CyTOF, 150 for flow cytometry, higher for spectral flow
  arcsinh_cofactor: 150
  # attempt to automatically separate non-mass channels for CyTOF data
  cytof: FALSE
# quarto render hdc-2-prepare-sce.qmd -P project_name:? -P arcsinh_cofactor:?
---

This script generates a SingleCellExperiment object with sample annotations.
The previous step (`hdc-1-import-fcs.qmd`) saves an unfiltered flowSet object.
It generates `input/metadata-files.csv` and `input/metadata-channels.csv` tables used to identify the relevant samples and markers/antibodies.
Those files should be edited to clean up labels, define sample groups, and exclude samples or markers/antibodies.
The primary comparison groups need to be defined in the `condition` column of `metadata-files.csv`.
Other metadata columns should be added as needed.
It's generally helpful for downstream analysis and visualization to simplify sample names (`sample_id` column in `metadata-files.csv`) and channel names (`antigen` column in `metadata-channels.csv`) by removing any unwanted prefixes or suffixes.
Problematic or unwanted samples and markers can be removed by deleting the corresponding rows.
The generated SingleCellExperiment object can then be used for CATALYST-based downstream analysis in the next step (`hdc-3-analyze-sce.qmd`).

# Settings

```{r packages}
#| message: false
#| warning: false
library(tidyverse)
library(glue)
library(cowplot)
library(qs2)
library(flowCore)
library(CATALYST)
```

Check parameters

```{r params}
params
```

Define inputs

```{r settings-inputs}
out_dir <- glue("./out-{params$project_name}")
input_dir <- glue("{out_dir}/input")
data_dir <- glue("{out_dir}/r-data")
qc_dir <- glue("{out_dir}/qc")
if (!dir.exists(out_dir)) stop("output dir does not exist: ", out_dir)

fs_qs2 <- glue("{data_dir}/flowset.qs2")
if (!file.exists(fs_qs2)) stop("flowSet does not exist: ", fs_qs2)

files_csv <- glue("{input_dir}/metadata-files.csv")
if (!file.exists(files_csv)) stop("files metadata table does not exist: ", files_csv)

channels_csv <- glue("{input_dir}/metadata-channels.csv")
if (!file.exists(channels_csv)) stop("channels metadata table does not exist: ", channels_csv)
```

Define outputs

```{r settings-outputs}
sce_qs2 <- glue("{data_dir}/sce.qs2")
```

# Import data

Import files metadata table

```{r import-files-csv}
files_tbl <- read_csv(files_csv, show_col_types = FALSE)
files_tbl
```

Validate files metadata table

```{r check-files-table}
if (!"file_name" %in% names(files_tbl)) {
  stop("files metadata table should have 'file_name' column")
}
if (!"sample_id" %in% names(files_tbl)) {
  stop("files metadata table should have 'sample_id' column (for CATALYST)")
}
if (!"patient_id" %in% names(files_tbl)) {
  stop("files metadata table should have 'patient_id' column (for CATALYST)")
}
if (!"condition" %in% names(files_tbl)) {
  stop("files metadata table should have 'condition' column (for CATALYST)")
}
if (anyNA(files_tbl$sample_id)) {
  stop("files metadata table 'sample_id' column has NAs")
}
if (anyNA(files_tbl$condition)) {
  stop("files metadata table 'condition' column has NAs")
}
if (n_distinct(files_tbl$condition) == 1) {
  stop("files metadata table should have multiple conditions")
}
if (!all(sapply(files_tbl, "class") %in% c("factor", "character"))) {
  stop("files metadata table column contents should be discrete")
}
```

Import channels metadata table

```{r import-channels-csv}
channels_tbl <- read_csv(channels_csv, show_col_types = FALSE)
channels_tbl
```

Validate channels metadata table

```{r check-channels-table}
if (!"fcs_colname" %in% names(channels_tbl)) {
  stop("channels metadata table should have 'fcs_colname' column")
}
if (!"antigen" %in% names(channels_tbl)) {
  stop("channels metadata table should have 'antigen' column (for CATALYST)")
}
```

Import flowSet

```{r import-flowSet}
fs <- qs2::qs_read(fs_qs2, validate_checksum = TRUE)
fs
```

Check that the parameter names are not all identical

```{r check-flowset-parameters}
if (n_distinct(fs[[1]]@parameters@data$name) == 1) stop("parameter names are all identical")
if (n_distinct(fs[[1]]@parameters@data$desc) == 1) stop("parameter descriptions are all identical")
```

Check if any files were removed from the files table

```{r}
removed_samples <- setdiff(rownames(pData(fs)), files_tbl$file_name)
removed_samples
```

Add removed files back to the table (will be removed after SingleCellExperiment is generated)

```{r}
if (length(removed_samples) > 0) {
  files_tbl <- bind_rows(files_tbl, data.frame(file_name = removed_samples))
}
```

```{r}
# remove rownames (causes an error with diffcyt)
# rownames(files_tbl) <- NULL
# rownames(channels_tbl) <- NULL
```

# Generate a SingleCellExperiment

Convert a flowSet into a SingleCellExperiment

```{r prepData}
arcsinh_cofactor <- params$arcsinh_cofactor
is_cytof <- as.logical(params$cytof)

# transform: arcsinh-transformation should be performed
# FACS: keep all channels as assay data
sce <-
  prepData(
    fs,
    panel = channels_tbl,
    md = files_tbl,
    transform = TRUE,
    cofactor = arcsinh_cofactor,
    FACS = !is_cytof
  )
sce
```

Remove files not found in the files table

```{r filter-samples}
sce <- filterSCE(sce, !is.na(sample_id))
sce
```

Check the contents of the SingleCellExperiment object

```{r}
rowData(sce)
```

```{r save-rowdata}
write_csv(as_tibble(rowData(sce), rownames = "feature_id"), glue("{qc_dir}/sce-rowdata.csv"))
```

`CATALYST::prepData()` does not keep all files metadata columns, so they need to be added

```{r clean-coldata}
colData(sce) <- colData(sce)[, "sample_id", drop = FALSE]

coldata_full_df <- left_join(as.data.frame(colData(sce)), files_tbl, by = c("sample_id"))
if (!all(coldata_full_df$sample_id == colData(sce)$sample_id)) stop("colData order mismatch")
coldata_full_df <- dplyr::select(coldata_full_df, !sample_id)
coldata_full_df <- dplyr::select(coldata_full_df, !file_name)
coldata_full_df <- dplyr::select(coldata_full_df, !full_path)

colData(sce) <- cbind(colData(sce), coldata_full_df)
colData(sce)
```

Add rownames to column metadata

```{r add-coldata-rownames}
colnames(sce) <- make.names(colData(sce)$sample_id, unique = TRUE)
```

Extract the experimental design table

```{r ei}
ei(sce)
```

```{r save-ei}
write_csv(as_tibble(ei(sce)), glue("{qc_dir}/sce-exp-design.csv"))
```

Plot the number of cells per sample

```{r plot-num-cells}
# plotCounts(sce, group_by = "sample_id", color_by = NULL)
# plotCounts(sce, group_by = "sample_id", color_by = "condition") + scale_fill_igv()
num_cells_plot <- plotCounts(sce, group_by = "sample_id", color_by = NULL)
ggsave(glue("{qc_dir}/sce-num-cells.png"), num_cells_plot, width = 15, height = 5)
```

Check expression

```{r check-assay-counts}
quantile(assay(sce, "counts")[, sample(1:ncol(sce), 1000)])
```

```{r check-assay-exprs}
quantile(assay(sce, "exprs")[, sample(1:ncol(sce), 1000)])
```

```{r}
# p = plotExprs(sce, color_by = "condition")
# p = plotExprs(sce, color_by = "sample_id")
# p$facet$params$ncol <- 6
# p
```

Save SingleCellExperiment

```{r save-sce}
qs2::qs_save(object = sce, file = sce_qs2)
```

Delete Rplots.pdf

```{r}
if (file.exists("Rplots.pdf")) {
  file.remove("Rplots.pdf")
}
```

# Session info

```{r session-info}
sessionInfo()
```
